Optimal (Multiway) Spatial Joins

### 2024-09-19 msj.tex
yufei

\def\vgap{\vspace{0mm}}
\def\extraspacing{\vspace{1mm} \noindent}
\def\figcapup{\vspace{-2mm}}
\def\figcapdown{\vspace{-4mm}}

@@

It is worth noting that $\Omega(n \log n)$ is a lower bound on the runtime of any comparison-based $k$-SJ algorithms (see Appendix~\ref{app:comp-class} for a formal definition of such algorithms), even for $k = 2$. This can be established via a reduction from the {\em element distinctness} problem; see \cite{dl79}.

%where we are given $n$ real values $e_1, e_2, ..., e_n$ and need to decide whether there are distinct $i, j \in [n]$ satisfying $e_i = e_j$. The ED problem demands $\Omega(n \log n)$ comparisons to solve \cite{dl79}.

@@

\section*{Appendix}

\section{Comparison-Based $k$-SJ Algorithms} \label{app:comp-class}

Fix an integer $k \ge 3$ and $k$ integers $n_1, n_2, ..., n_k$ at least 1. Define $n = \sum_{i=1}^k n_i$. An {\em $(n_1, n_2, ..., n_k)$-input} to the $k$-SJ problem comprises $k$ sets of rectangles $R_1, R_2, ..., R_k$ in $\real^2$ such that $|R_i| = n_i$ for each $i \in [k]$. We assume that each rectangle in $R_i$ ($i \in [k]$) is given a distinct ID from $[n_i]$.

A {\em comparison-based $k$-SJ} algorithm for solving $(n_1, n_2, ..., n_k)$-inputs is a tree $\T$ defined as follows. Each internal node $u$ of $\T$ has three child nodes, referred as the {\em left}, {\em middle}, and {\em right} child, respectively. Each node is associated with an {\em instruction} of the form $(i, \lambda, \side; i', \lambda', \side')$ where
\myitems{
    \item $i, i' \in [1, k]$;
    \item $\lambda \in [n_i], \lambda' \in [n_{i'}]$;
    \item $\side$ and $\side'$ are functions chosen from the set $\{\xleft, \xright, \ybot$, $\ytop\}$.
}
Each leaf is associated with an {\em result}, which is defined to be a subset of $[n_1] \times [n_2] \times ... \times [n_k]$.

Given an $(n_1, n_2, ..., n_k)$-input $\set{R_1, R_2, ..., R_k}$, we execute the algorithm by traversing a root-to-leaf path in $\T$. In general, at an internal node with the instruction $(i, \lambda, \side; i', \lambda', \side')$, we first identify
\myitems{
    \item $r$ as the rectangle with ID $\lambda$ in $R_i$;
    \item $r'$ as the rectangle with ID $\lambda'$ in $R_{i'}$.
}
Then, we compare $\side(r)$ to $\side'(r')$ (e.g., if $\side = \xright$ and $\side' = \ybot$, then the comparison is between $\xright(r)$ to $\ybot(r')$). We branch to
\myitems{
    \item the left child if $\side(r) < \side'(r')$;
    \item the middle child if $\side(r) = \side'(r')$;
    \item the right child if $\side(r) > \side'(r')$.
}
At a leaf, we look at its associated result and, for each $(\lambda_1, \lambda_2, ..., \lambda_k)$ in that result, output $(r_1, r_2, ..., r_k)$ where $r_i$ ($i \in [k]$) is the rectangle with ID $\lambda_i$ in $R_i$. The algorithm is required to produce the correct output on all $(n_1, n_2, ..., n_k)$-inputs. Its {\em worst-case cost} is the height of $\T$.

Through a reduction from element distinctness \cite{dl79}, one can show that, for $k = 2$, any comparison-based algorithm correctly solving all $(n/2, n/2)$-inputs must have a worst-case cost of $\Omega(n \log n)$.

@@

\extraspacing {\bf Algorithm for Problem $\bm{\mathscr{B}}$.} Sort the set of y-coordinates defined by all the segments of $H$ and $V$. Next, sweep (conceptually) a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, maintain the set $V_\ell$ of segments in $V$ that intersect with $\ell$. Let $S_\ell$ be the set of x-coordinates of the segments in $V_\ell$; we store $S_\ell$ in a binary search tree (BST) $\T$. Specifically, when $\ell$ hits the lower  (resp., upper) endpoint of a vertical segment $v = x \times [y_1, y_2]$ of $V$, we insert (resp., delete) the value $x$ into (resp., from) $\T$, which can be done in $O(\log n)$ amortized time. When $\ell$ hits a horizontal segment $h = [x_1, x_2] \times y$ of $H$, search $\T$ to determine the successor $x'$ of $x_1$ in $S_\ell$. If $x' \le x_2$, then we output a pair $(h, p)$, where $p$ is the point $(x', y)$; otherwise, output nothing for $h$. The overall running time is $O(n \log n)$.

\extraspacing {\bf Algorithm for Problem $\bm{\mathscr{C}}$.} Sort the set of y-coordinates defined by all the segments in $H$ and all the rectangles in $R$. Next, sweep (conceptually) a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, maintain the set $R_\ell$ of rectangles in $R$ that intersect with $\ell$. Let $S_\ell$ be the set of x-ranges of the rectangles in $R_\ell$; we store $S_\ell$ in a stabbing-max structure $\T$ of \cite{aak+12}. Specifically, when $\ell$ hits the bottom (resp., top) edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we insert (resp., delete) $[x_1, x_2]$ with {\em weight} $x_2$ into (resp., from) $\T$, which can be done in $O(\log n)$ time. When $\ell$ hits a horizontal segment $h = [x_1, x_2] \times y$ of $P$, search $\T$ to determine the maximum weight $w$ of all the intervals in $S_\ell$ containing the value $x_1$. If $w \ne -\infty$, we output $(h, p)$ where the point $p$ is defined in a way depending on $w$: if $w \le x_2$, then $p = (w, y)$; otherwise $p = (x_2, y)$. The overall running time is $O(n \log n)$.

@@

Sweep (conceptually) a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, we maintain the set $R_\ell$ of {\em active} rectangles in $R$ that intersect with $\ell$. Let $S_\ell$ be the set of x-ranges of the rectangles in $R_\ell$; we store $S_\ell$ in a PST $\T$. Specifically, when $\ell$ hits the top edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we insert $[x_1, x_2]$ into $\T$ and mark $r$ as active, which can be done in $O(\log n)$ time. The rectangle $r$ will be referred to as the {\em host} of $[x_1, x_2]$ and is stored together with $[x_1, x_2]$ in $\T$.

@@

Next, we aim to produce, for each pair $(h, r) \in H \times r$ such that $h$ crosses $r$, a pair $(h, \lambda)$ where $\lambda$ is the ID of $r$. These pairs may be output in an arbitrary order. Sort the set of y-coordinates defined by all the horizontal segments and rectangles. Sweep (conceptually) a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, we maintain the set $R_\ell$ of rectangles in $R$ that intersect with $\ell$. Let $S_\ell$ be the set of x-ranges of the rectangles in $R_\ell$; we store $S_\ell$ in a PST $\T$. Specifically, when $\ell$ hits the bottom (resp., top) edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we insert (resp., delete) $[x_1, x_2]$ into (resp., from) $\T$, which can be done in $O(\log n)$ time. We call the rectangle $r$ the {\em host} of $[x_1, x_2]$ and store its ID together with $[x_1, x_2]$ in $\T$. When $\ell$ hits a horizontal segment $h = [z_1, z_2] \times y$ of $H$, perform a reverse-containment query on $\T$ to find all the intervals in $S_\ell$ that contain $[z_1, z_2]$; the query cost is $O(\log n + K_h)$, where $K_h$ is the number of intervals reported. For each retrieved interval $[x_1, x_2]$, we also obtain the ID $\lambda$ of its host rectangle $r$, and output the pair $(h, \lambda)$.

@@

As mentioned in Section~\ref{sec:intro:prev}, 2-SJ can be solved in $F_2(n, \out) = O(n \log n + \out)$ time using a comparison-based algorithm. By Theorem~\ref{thm:main-recur}, in general, a comparison-based $(k-1)$-SJ algorithm with runtime $F_{k-1}(n, \out)$ spawns a comparison-based $k$-SJ algorithm whose running time $F_k(n, \out)$ obeys \eqref{eqn:main:reccurrence}. Specifically, for $k \ge 3$, there is a constant $c \ge 2$ such that
\myeqn{
    && F_{k}(n, \out) \nn \\
    &\le& c \cdot k^3 \cdot \big( F_{k-1}(n, \out) + n \log n + k \cdot \out \big) \nn \\
    &=& c^2 \cdot k^3(k-1)^3 \cdot (F_{k-2}(n,\out) + n\log n + k\cdot \out) + c \cdot k^3 \cdot (n\log n + k\cdot \out ) \nn\\
    &<& c^2 \cdot k^3(k-1)^3 \cdot F_{k-2}(n,\out) +  (c+c^2)\cdot k^3(k-1)^3 \cdot (n\log n + k \cdot\out) \nn \\
    &\leq& c^3 \cdot k^3(k-1)^3(k-2)^3 \cdot F_{k-3}(n,\out) +
    (c+c^2+c^3)\cdot k^3(k-1)^3(k-2)^3 \cdot (n\log n + k \cdot\out) \nn\\
    &\leq& ... \nn \\
    &\leq& c^{k-2} \cdot (k!)^3 \cdot F_{2}(n,\out) + \Big(\sum_{i =1}^{k-2}c^i \Big)\cdot (k!)^3 \cdot (n\log n + k \cdot\out) \nn \\
    &\leq & 2c^{k-1} \cdot (k!)^3 \cdot O(n\log n + k \cdot\out) \nn
}
which completes the proof of Theorem~\ref{thm:main-alg}.

>>>>

\def\vgap{\vspace{0mm}}
\def\extraspacing{\vspace{1mm} \noindent}
\def\figcapup{\vspace{-2mm}}
\def\figcapdown{\vspace{-5mm}}

@@

It is worth noting that $\Omega(n \log n)$ is a lower bound on the runtime of any comparison-based $k$-SJ algorithms
%(see Appendix~\ref{app:comp-class} for a formal definition of such algorithms),
even for $k = 2$. This can be established via a reduction from the {\em element distinctness} problem; see \cite{dl79}.

@@

\section*{Appendix}
%
% \section{Comparison-Based $k$-SJ Algorithms} \label{app:comp-class}
%
% Fix an integer $k \ge 3$ and $k$ integers $n_1, n_2, ..., n_k$ at least 1. Define $n = \sum_{i=1}^k n_i$. An {\em $(n_1, n_2, ..., n_k)$-input} to the $k$-SJ problem comprises $k$ sets of rectangles $R_1, R_2, ..., R_k$ in $\real^2$ such that $|R_i| = n_i$ for each $i \in [k]$. We assume that each rectangle in $R_i$ ($i \in [k]$) is given a distinct ID from $[n_i]$.
%
% \vgap
%
% A {\em comparison-based $k$-SJ} algorithm for solving $(n_1, n_2, ..., n_k)$-inputs is a tree $\T$ defined as follows. Each internal node $u$ of $\T$ has three child nodes, referred as the {\em left}, {\em middle}, and {\em right} child, respectively. Each node is associated with an {\em instruction} of the form $(i, \lambda, \side; i', \lambda', \side')$ where
% \myitems{
%     \item $i, i' \in [1, k]$;
%     \item $\lambda \in [n_i], \lambda' \in [n_{i'}]$;
%     \item $\side$ and $\side'$ are functions chosen from the set $\{\xleft, \xright, \ybot$, $\ytop\}$.
% }
% Each leaf is associated with an {\em result}, which is defined to be a subset of $[n_1] \times [n_2] \times ... \times [n_k]$.
%
% \vgap
%
% Given an $(n_1, n_2, ..., n_k)$-input $\set{R_1, R_2, ..., R_k}$, we execute the algorithm by traversing a root-to-leaf path in $\T$. In general, at an internal node with the instruction $(i, \lambda, \side; i', \lambda', \side')$, we first identify
% \myitems{
%     \item $r$ as the rectangle with ID $\lambda$ in $R_i$;
%     \item $r'$ as the rectangle with ID $\lambda'$ in $R_{i'}$.
% }
% Then, we compare $\side(r)$ to $\side'(r')$ (e.g., if $\side = \xright$ and $\side' = \ybot$, then the comparison is between $\xright(r)$ to $\ybot(r')$). We branch to
% \myitems{
%     \item the left child if $\side(r) < \side'(r')$;
%     \item the middle child if $\side(r) = \side'(r')$;
%     \item the right child if $\side(r) > \side'(r')$.
% }
% At a leaf, we look at its associated result and, for each $(\lambda_1, \lambda_2, ..., \lambda_k)$ in that result, output $(r_1, r_2, ..., r_k)$ where $r_i$ ($i \in [k]$) is the rectangle with ID $\lambda_i$ in $R_i$. The algorithm is required to produce the correct output on all $(n_1, n_2, ..., n_k)$-inputs. Its {\em worst-case cost} is the height of $\T$.
%
%
% \vgap
%
% Through a reduction from element distinctness \cite{dl79}, one can show that, for $k = 2$, any comparison-based algorithm correctly solving all $(n/2, n/2)$-inputs must have a worst-case cost of $\Omega(n \log n)$.

@@

\extraspacing {\bf Algorithm for Problem $\bm{\mathscr{B}}$.} Sort the set of y-coordinates defined by all the segments of $H$ and $V$. Next, sweep a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, maintain the set $V_\ell$ of segments in $V$ that intersect with $\ell$. Let $S_\ell$ be the set of x-coordinates of the segments in $V_\ell$; we store $S_\ell$ in a binary search tree (BST) $\T$. Specifically, when $\ell$ hits the lower  (resp., upper) endpoint of a vertical segment $v = x \times [y_1, y_2]$ of $V$, we insert (resp., delete) the value $x$ into (resp., from) $\T$, which can be done in $O(\log n)$ amortized time. When $\ell$ hits a horizontal segment $h = [x_1, x_2] \times y$ of $H$, search $\T$ to determine the successor $x'$ of $x_1$ in $S_\ell$. If $x' \le x_2$, then we output a pair $(h, p)$, where $p$ is the point $(x', y)$; otherwise, output nothing for $h$. The overall running time is $O(n \log n)$.

\extraspacing {\bf Algorithm for Problem $\bm{\mathscr{C}}$.} Sort the set of y-coordinates defined by all the segments in $H$ and all the rectangles in $R$. Next, sweep a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, maintain the set $R_\ell$ of rectangles in $R$ that intersect with $\ell$. Let $S_\ell$ be the set of x-ranges of the rectangles in $R_\ell$; we store $S_\ell$ in a stabbing-max structure $\T$ of \cite{aak+12}. Specifically, when $\ell$ hits the bottom (resp., top) edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we insert (resp., delete) $[x_1, x_2]$ with {\em weight} $x_2$ into (resp., from) $\T$, which can be done in $O(\log n)$ time. When $\ell$ hits a horizontal segment $h = [x_1, x_2] \times y$ of $P$, search $\T$ to determine the maximum weight $w$ of all the intervals in $S_\ell$ containing the value $x_1$. If $w \ne -\infty$, we output $(h, p)$ where the point $p$ is defined in a way depending on $w$: if $w \le x_2$, then $p = (w, y)$; otherwise $p = (x_2, y)$. The overall running time is $O(n \log n)$.

@@

Sweep a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, we maintain the set $R_\ell$ of {\em active} rectangles in $R$ that intersect with $\ell$. Let $S_\ell$ be the set of x-ranges of the rectangles in $R_\ell$; we store $S_\ell$ in a PST $\T$. Specifically, when $\ell$ hits the top edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we insert $[x_1, x_2]$ into $\T$ and mark $r$ as active, which can be done in $O(\log n)$ time. The rectangle $r$ will be referred to as the {\em host} of $[x_1, x_2]$ and is stored together with $[x_1, x_2]$ in $\T$.

@@

Next, we aim to produce, for each pair $(h, r) \in H \times r$ such that $h$ crosses $r$, a pair $(h, \lambda)$ where $\lambda$ is the ID of $r$. These pairs may be output in an arbitrary order. Sort the set of y-coordinates defined by all the horizontal segments and rectangles. Sweep a horizontal line $\ell$ from $y = -\infty$ to $y = \infty$. At all times, we maintain the set $R_\ell$ of rectangles in $R$ that intersect with $\ell$. Let $S_\ell$ be the set of x-ranges of the rectangles in $R_\ell$; we store $S_\ell$ in a PST $\T$. Specifically, when $\ell$ hits the bottom (resp., top) edge of a rectangle $r = [x_1, x_2] \times [y_1, y_2]$ of $R$, we insert (resp., delete) $[x_1, x_2]$ into (resp., from) $\T$, which can be done in $O(\log n)$ time. We call the rectangle $r$ the {\em host} of $[x_1, x_2]$ and store its ID together with $[x_1, x_2]$ in $\T$. When $\ell$ hits a horizontal segment $h = [z_1, z_2] \times y$ of $H$, perform a reverse-containment query on $\T$ to find all the intervals in $S_\ell$ that contain $[z_1, z_2]$; the query cost is $O(\log n + K_h)$, where $K_h$ is the number of intervals reported. For each retrieved interval $[x_1, x_2]$, we also obtain the ID $\lambda$ of its host rectangle $r$, and output the pair $(h, \lambda)$.

@@

As mentioned in Section~\ref{sec:intro:prev}, 2-SJ can be solved in $F_2(n, \out) = O(n \log n + \out)$ time using a comparison-based algorithm. By Theorem~\ref{thm:main-recur}, in general, a comparison-based $(k-1)$-SJ algorithm with runtime $F_{k-1}(n, \out)$ spawns a comparison-based $k$-SJ algorithm whose running time $F_k(n, \out)$ obeys \eqref{eqn:main:reccurrence}. Specifically, for $k \ge 3$, there is a constant $c \ge 2$ such that
\myeqn{
    && F_{k}(n, \out) \nn \\
    &\le& c \cdot k^3 \cdot \big( F_{k-1}(n, \out) + n \log n + k \cdot \out \big) \nn \\
    &=& c^2 \cdot k^3(k-1)^3 \cdot (F_{k-2}(n,\out) + n\log n + k\cdot \out) + c \cdot k^3 \cdot (n\log n + k\cdot \out ) \nn\\
    &<& c^2 \cdot k^3(k-1)^3 \cdot F_{k-2}(n,\out) +  (c+c^2)\cdot k^3(k-1)^3 \cdot (n\log n + k \cdot\out) \nn \\
    &\leq& c^3 \cdot k^3(k-1)^3(k-2)^3 \cdot F_{k-3}(n,\out) +
    (c+c^2+c^3)\cdot k^3(k-1)^3(k-2)^3 \cdot (n\log n + k \cdot\out) \nn\\
    &\leq ... \leq & %\nn \\
    c^{k-2} \cdot (k!)^3 \cdot F_{2}(n,\out) + \Big(\sum_{i =1}^{k-2}c^i \Big)\cdot (k!)^3 \cdot (n\log n + k \cdot\out) \nn \\
    &\leq & 2c^{k-1} \cdot (k!)^3 \cdot O(n\log n + k \cdot\out) \nn
}
which completes the proof of Theorem~\ref{thm:main-alg}.
